---
title: 'Measuring Euro Area monetary policy: going beyond monetary surprises'
subtitle: by Umberto Collodel & Vanessa Kunzmann
output:
  html_document:
    df_print: paged
  pdf_document:
    latex_engine: xelatex
editor_options:
  markdown:
    wrap: 72
---

# Theoretical Foundation

Figure 1 shows a stylized comparison between a monetary surprise a-la
Altavilla and a combined monetary and monetary policy uncertainty
surprise.

In the left panel, before the governing council,the distribution of OIS
quotes is centered around 0 and has a unit standard deviation. After the
meeting, the distribution shifts to the right, centering at 1; this
movement coincides with a tightening of the mean OIS rate.

In the right panel, instead, not only the mean of the distribution
shifts to right after the governing council, but its variance also
increases. Although agents revised their expectation, there is more
heterogeneity concerning the future path of the rate, with different
agents betting on different movements of the aforementioned.

The two cases could have very different implications for the behaviour
of financial variables (e.g. stock prices, exchange rates, etc.) and
trasmission of monetary policy (sovereign yields), and thus for real
variables such as credit, output and inflation. In principle, one would
expect uncertainty to hamper the effect of monetary surprises, hence
hindering ECB decided policies.

```{r setup, include=FALSE}
library(readxl)
library(tidyverse)
library(showtext)
library(zoo)
library(ggtext)
library(extrafont)
font_add("Segoe UI Light", "C:/Windows/Fonts/segoeuil.ttf")
showtext_auto()


knitr::opts_chunk$set(echo = FALSE, message = F, warning = F)



```


```{r,fig.width=12}

set.seed(16)

# Create data for the first facet
data1 <- data.frame(
  x = c(rnorm(1000, mean = 0, sd = 1), rnorm(1000, mean = 1, sd = 2)),
  group = factor(rep(c("Mean 0, SD 1", "Mean 1, SD 1.5"), each = 1000))
)

# Create data for the second facet
data2 <- data.frame(
  x = c(rnorm(1000, mean = 0, sd = 1), rnorm(1000, mean = 1, sd = 1)),
  group = factor(rep(c("Mean 0, SD 1", "Mean 1, SD 1.5"), each = 1000))
)

# Combine data and add a facet identifier
data1$facet <- "Monetary and Uncertainty Surprise"
data2$facet <- "Monetary Surprise"
data <- rbind(data1, data2)

# Plot the data
data %>% 
mutate(facet = factor(facet, levels = c("Monetary Surprise",
                                        "Monetary and Uncertainty Surprise"))) %>% 
ggplot(aes(x = x, fill = group)) +
  geom_density(alpha = 0.3, col="white") +
  facet_wrap(~facet) +
  theme_minimal() +
    labs(title="Figure 1: Stylized example of combined monetary and monetary uncertainty surprise",
       x="",
       y="",
       col="") +
  theme_bw() +
  theme(text=element_text(family="Segoe UI Light")) +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank()) +
  theme(legend.position = "none",
        legend.text = element_text(size=14)) +
  theme(
    plot.title = element_text(
      size = 22,            # Set font size
      face = "bold", # Make the title bold and italic
      color = "black",
      family = "Segoe UI Light")) + 
    theme( axis.text = element_text( size = 14 ),
         axis.text.x = element_text( size = 16 ),
         axis.title = element_text( size = 16 ),
         strip.text = element_text(size = 16)) +
  scale_fill_manual(values=c("blue", "red"))


```

# Data

We retrieve daily data on OIS euro rates at different tenors
(3m,6m,1Y,2Y,5Y and 10Y) from 2001 up to today. However, long-term rates
are only available from 2011 onwards, so we decide to work only on this
part of the sample (maybe use German yields before?). For each day, we
have available the high and low quotes, and opening and closing.

To proxy the standard deviation before and after governing councils, we
use the daily range (high minus low). ADD LITERATURE We then average the
daily range over the three days before and after, to avoid sensitivity
to extreme observations.

To provide an llustration of the (cleansed) daily data and of monetary
policy uncertainty surprises, Fig. 2 shows the change in the 10Y OIS
rate (from the EA-MPD database) around governing councils, and the
constructed standard deviation, before and after. We show four different
policy meeting dates: 7 November 2013, 3 December 2015, 7 October 2017,
and 25 Luglio 2019. We select the 10-year rate because changes in the
varianc of the distribution are more likely to occur at long horizons.

```{r, fig.width=10, fig.height=6}

dates_event=as.Date(c("2013-11-07","2015-12-03","2017-09-07",
                      "2019-07-25"))

dates_before_after <- dates_event %>% 
  map(~ seq(from = .x - 4, to = .x + 5, by = "day")) %>% 
  unlist() %>% 
  as.Date(., origin = "1970-01-01")

dates_before <- dates_event %>% 
  map(~ seq(from = .x - 4, to = .x -1, by = "day")) %>% 
  unlist() %>% 
  as.Date(., origin = "1970-01-01")

dates_after <- dates_event %>% 
  map(~ seq(from = .x + 1, to = .x + 5, by = "day")) %>% 
  unlist() %>% 
  as.Date(., origin = "1970-01-01")




 event_study_df <-read_excel("../raw_data/OIS.xls", sheet = "10Y_bid") %>%
  slice(-1) %>% 
  select(1:5) %>% 
  mutate(Daily = as.numeric(Daily)) %>% 
  mutate(Daily = as.Date(Daily, origin = "1899-12-30")) %>% 
  setNames(c("daily","first","high","low","last")) %>% 
  mutate_at(2:ncol(.),as.numeric) %>% 
  filter(daily %in% dates_before_after) %>%
  mutate(year = year(daily)) %>% 
  mutate(before_govc = case_when(daily %in% dates_before ~1,
                                 T ~ 0),
         after_govc = case_when(daily %in% dates_after ~ 1,
                                T ~ 0)) %>% 
  group_by(year, before_govc) %>% 
  mutate(mean_before = mean(high-low)*100) %>% 
  ungroup() %>% 
  group_by(year, after_govc) %>% 
  mutate(mean_after = mean(high-low)*100) %>% 
  ungroup() %>% 
  mutate(correct_mean_after = lead(mean_after,1),
         correct_mean_before = lag(mean_before,1)) %>% 
  filter(daily %in% dates_event)

   
mean_rate_df <- read_xlsx("../../PEPP_effect/raw_data/00EA_MPD_update_january2024.xlsx", 
          sheet = "Monetary Event Window") %>% 
  select(date,OIS_10Y,STOXX50) %>%
  mutate(date = as.Date(date)) %>% 
  filter(date %in% dates_event) %>% 
  setNames(c("daily","mean_change","stock_change"))
  

 
event_study_df %>%
  inner_join(mean_rate_df) %>% 
  pivot_longer(correct_mean_after:mean_change,names_to = "time_id",values_to = "value") %>% 
  mutate(time_id = case_when(time_id == "correct_mean_before" ~ "Before",
                             time_id == "correct_mean_after" ~ "After",
                             time_id == "mean_change" ~ "Rate Δ")) %>% 
  mutate(time_id = factor(time_id,levels = c("Rate Δ","Before","After"))) %>% 
  ggplot(aes(time_id,value,fill=time_id)) +
  geom_col(width=0.2,alpha=0.8,col="black") +
  geom_segment(aes(x = 1.5, xend = 1.5, y = -Inf, yend = Inf), color = "black") + 
  geom_hline(yintercept = 0) +
  scale_fill_manual(values = c("#F8766D","#434343","#818181")) +
  labs(title="Figure 2: Daily 10-year OIS rate and range change before and after governing councils",
       y="Basis Points",
       x="",
       fill="",
       caption = "**Note**: Before and after ranges are calculated as the difference between high and
       low quote: we report three-days averages before and after the day of the governing council.
       **Source**: Thomson Reuters") +
  facet_wrap(~ daily, scales = "free") +
  theme_bw() +
   theme(text=element_text(family="Segoe UI Light")) +
  theme(legend.position = "none") + # Remove legend
  theme(
    plot.title = element_text(
      size = 18,            # Set font size
      face = "bold", # Make the title bold and italic
      color = "black",
      family = "Segoe UI Light")) + # Set the font color    )) +
  theme( axis.text = element_text( size = 14 ),
         axis.text.x = element_text( size = 16 ),
         axis.title = element_text( size = 16 ),
         strip.text = element_text(size = 18)) +
  theme(panel.grid.major.x = element_blank()) +
  theme(plot.caption = element_text(hjust = 0,size=12))
  


  
```

The first panel shows the movement in the 10Y OIS following the
2013-11-07 Governing Council: there a was a negative monetary surprise,
a loosening of the rate by close to 5bps, and a widening in the range,
from 4 to 8 bps.This date coincides with one of the first meetings in
which the ECB introduced forward guidance on the future path of interest
rates. The panel 2015-12-03 depicts a policy date in which there was a
sizeable positive monetary surprise for the 10Y OIS. This episode
captures the financial markets' disappointment following the ECB
decision to increase the size of its QE program: markets evidently were
expecting a larger increase of QE. Like the previous episode, the range
of quotes roughly doubles, going from 5bps to more than 10bps.The third
panel shows a day in which there was no monetary surprise and the range
basically does not vary. Policy dates like these are surprisingly rare;
there is usually some news for the financial markets after the whole
monetary event.

```{r}
#Add table with factors significance and variance share

fit[[2]]



```

```{r}
loadings_final_df %>% 
  ggplot(aes(Term,value,group=1)) +
  geom_line(size=1.5,col="red") +
  geom_point(size=4,alpha=0.3,col="red") +
  facet_wrap(~Factor) +
  labs(title="Figure 3: Factor Loadings",
       col="",
       y="",
       x="",
       fill="",
       caption = "**Note**: The model decomposes movements in OIS (1m, 3m, 6m, 1y, 2y, 5y, 10y) into 
       three policy factors: downward-sloping (Timing), hump-shaped (Path) and upward-sloping (QE).  
       It takes three principal components from the variation in yields range around GovC meetings,
       rotates and scales them.  
**Source**: Authors' calculation   
**Latest observation**: 15 June 2023.") +  
  theme_bw() +
  theme(text=element_text(family="Segoe UI Light")) +
  theme(plot.caption = element_text(hjust=0)) +
  theme(axis.text.x = element_text(vjust = 0.5, hjust=0.5)) +
  theme( axis.text = element_text( size = 14 ),
         axis.text.x = element_text( size = 20 ),
         axis.title = element_text( size = 16, face = "bold" ),
         legend.text = element_text(size=14),
         # The new stuff
         strip.text = element_text(size = 20)) +
  theme(
    plot.title = element_text(
      size = 20,            # Set font size
      face = "bold", # Make the title bold and italic
      color = "black",
      family = "Segoe UI Light")) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  theme(plot.caption = element_text(hjust = 0,size=12))
```

```{r}
# Add indvidual time series scaled and winsorized.
```

```{r}
time_serie_df %>% 
  filter(year(date) >= 2015) %>% 
  ggplot(aes(date,value, fill=Factor)) +
  geom_col(width = 17) +
  labs(title="Figure 4: Monetary Uncertainty Surprises over Time",
       y="Sd",
       x="",
       fill="",
       caption = "**Note**: The model decomposes movements in OIS (1m, 3m, 6m, 1y, 2y, 5y, 10y) into three policy factors: downward-sloping (Timing), hump-shaped (Path) and upward-sloping (QE). It takes three principal components  
from the variation in yields around GovC meetings (Press Release and Press Conference), rotates and scales them. An additional model employs the same methodology for the variation of 10 years spreads (IT-DE, SP-DE   
and FR-DE) against German bunds around GovC meetings. The movements can be interpreted as similar to a 1 standard deviation movement in the reference asset for that factor (OIS 1m, OIS 1y, OIS 10y, IT-DE 10Y Spread).  
**Source**: Authors' calculation, Bloomberg, EA-MPD (Altavilla et al., 2019)  
**Latest observation**: 15 June 2023.") +  
  theme_bw() +
  theme(
    plot.title = element_text(
      size = 20,            # Set font size
      face = "bold", # Make the title bold and italic
      color = "black",
      family = "Segoe UI Light")) +
  theme(text=element_text(family="Segoe UI Light")) +
  theme(plot.caption = element_text(hjust=0)) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  scale_fill_manual(values=c("#35BBCA","#0191B4","#D3DD18","#FE7A15")) +
  theme( axis.text = element_text( size = 14 ),
         axis.text.x = element_text( size = 20 ),
         axis.title = element_text( size = 16 ),
         legend.position="bottom",
         legend.text = element_text(size=14),
         strip.text = element_text(size = 20)) +
  theme(plot.caption = element_text(hjust = 0,size=12))

```

```{r, eval=FALSE}
# (overall correlation)


# Get dates of govc

width=23

dates_cor=time_serie_df %>% 
  select(date) %>%
  filter(date <= "2023-06-15") %>% 
  unique() %>% 
  slice(-c(1:width))

# Plot:

df %>% 
  pivot_wider(names_from = moment, values_from = value) %>% 
  select(mon,uncertainty) %>% 
  mutate(mon = abs(mon)) %>% 
  rollapply(width = 24*3,  # window size
            FUN = function(x) cor(x),
            by.column = FALSE) %>% 
  as_tibble() %>% 
  select(mon2) %>% 
  rowid_to_column("date") %>%
  mutate(date = rep(dates_cor %>% pull(date),each = 3) %>% head(-2)) %>%
  # Work on the timeline labels
  mutate(label = case_when(date == "2017-10-26" ~ "Zero Lower Bound",
                           date == " 2023-02-02" ~ "Tightening Cycle",
                           date == "2021-07-22" ~ "Covid-19 Pandemic",
                           T ~ NA)) %>% 
  mutate(y_label = case_when(date == "2017-10-26" ~ -0.2,
                           date == " 2023-02-02" ~ -0.3,
                           date == "2021-07-22" ~ 0.3,
                           T ~ NA)) %>% 
  mutate(date_label = case_when(date == "2017-10-26" ~ date,
                             date == " 2023-02-02" ~ date,
                             date == "2021-07-22" ~ date,
                             T ~ NA)) %>% 
  mutate(yend_label = case_when(date == "2017-10-26" ~ mon2-0.02,
                                date == " 2023-02-02" ~ mon2 -0.02,
                                date == "2021-07-22" ~ mon2 +0.02,
                                T ~ NA)) %>%
  ###
  ggplot(aes(date,mon2,group=1)) +
  geom_line(size=1.5, col="orange") +
  geom_smooth(method = "loess", span = 0.2, color = "red") +
  geom_hline(yintercept = 0, size =1, linetype = "dashed") +
  # Work on the timeline labels
  geom_segment(aes(x = date_label,y = y_label,xend = date_label, yend =yend_label),size=0.5) +
  geom_text(aes(x = date_label,y = y_label,label = label),hjust = 1.1,vjust = 1.1,parse = FALSE) +
  geom_point(aes(x = date_label,y = y_label),size=2,alpha=0.5) +
  ###
  labs(title="Correlation - Monetary and monetary uncertainty surprises",
       subtitle = expression(italic("2yrs rolling window")),
       col="",
       y="",
       x="",
       fill="",
       caption = "**Note**: 
             **Source**: Authors' calculation   
             **Latest observation**: 15 June 2023.") +
  ylim(-0.4,0.4) +
  theme_bw() +
  theme(plot.caption = element_text(hjust=0)) +
  theme(title = element_text(size=20)) +
  theme(axis.text.x = element_text(vjust = 0.5, hjust=0.5)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(axis.text = element_text( size = 14 ),
        axis.text.x = element_text( size = 20 ),
        axis.title = element_text( size = 16, face = "bold" ),
        legend.text = element_text(size=14),
        # The new stuff
        strip.text = element_text(size = 20)) +
  theme(plot.caption = element_markdown(hjust = 0,size=12))


```
